HyperLogLog: Exponentially Bad in Adversarial Settings

### 2021-09-07 docs/HLL_exp_bad/main.tex
rejigging security analysis once again

\begin{figure*}[tb!]
\centering
     \makebox[\linewidth]{
\fbox{
\begin{minipage}{40em}
\begin{flushleft}
\textit{Let $h : \mathcal{D} \rightarrow \{0, 1\}^{32}$ hash data from some domain $\mathcal{D}$ to binary $32$-bit strings.\\
For $s$ of bit-length $\ell$, let $\rho(s)$ denote the position of the leftmost $1$-bit of $s$: e.g., $\rho(1...) = 1, \rho(0001...) = 4, \rho(0^\ell) = \ell + 1.$\\}
\textbf{define} $m = 2^n$ with $n \in [4..16].$\\
\textbf{define} $\alpha_{16} = 0.673$; $\alpha_{32} = 0.697$; $\alpha_{64} = 0.709$; $\alpha_{m} = 0.7213/(1+ 1.079/m)$ for $m \ge 128$;\\
\vspace{0.5em}
\textbf{Algorithm} HyperLogLog (\textbf{input} ${\mathcal{X}}$ : multiset of items from domain $\mathcal{D}$).\\
\textbf{initialize} a collection of $m$ registers, $M[1], ..., M[m]$, to $0$;\\
\vspace{0.5em}
\textbf{for} $x \in {\mathcal{X}}$ \textbf{do}\\
    \hspace{2em} \textbf{set} $y := h(x);$\\
    \hspace{2em} \textbf{set} $j = 1 + <y_1y_2...y_{n}>_2$; \hfill \textit{\{bucket index determined by the first $n$ bits of $y$\}}\\
    \hspace{2em} \textbf{set} $w := y_{n+1}y_{n+2} \ldots y_{n+\ell}$;\hfill \textit{\{bit string determined by the last $\ell$ bits of $y$\}}\\
   \hspace{2em} \textbf{set} $M[j] :=$ max$(M[j], \rho(w))$;\\
\vspace{0.2em}
\textbf{compute} $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$; \hfill \textit{\{the raw HyperLogLog estimate\}}\\
\vspace{0.2em}
\textbf{if} $E \le \frac{5}{2}m$ \textbf{then}\\
    \hspace{2em}  \textbf{let} $V$ be the number of registers equal to $0$;\\
    \hspace{2em}  \textbf{if} $V \neq 0$ \textbf{then set} $E^*:=  m\cdot \ln(m/V)$ \textbf{else set} $E^*:= E$; \hfill \textit{\{small range correction\}}\\
\textbf{if} $E \le \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em} \textbf{set} $E^*:= E$; \hfill \textit{\{intermediate range --- no correction\}}\\
\textbf{if} $E > \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em}  \textbf{set} $E^*:= -2^{32}$ln$(1 - E/2^{32})$; \hfill \textit{\{large range correction\}}\\
\textbf{return} \textit{cardinality estimate $E^*$ with typical relative error} $\pm1.04/\sqrt{m}$.
\end{flushleft}
\end{minipage}
}}
\caption{The HyperLogLog Algorithm from~\cite{hll}}
\label{fig:hll}
\end{figure*}

@@

In the original HLL proposal~\cite{hll}, relatively small and large cardinalities are handled separately to correct for systematic errors. When the raw cardinality estimate is below $\frac{5}{2}m$, the Hit Counting algorithm of Whang \emph{et al.}~\cite{hitcounting} is used to produce the final cardinality estimate. This estimator uses the number of empty buckets (whose register value is 0) instead of the number of leading zeros in the stream. A similar correction is applied for large cardinalities in order to take into account the likelihood of hash collisions. In addition, the cardinality estimate is usually rounded to a whole number before being output. The original HLL algorithm from~\cite{hll} is shown in Figure~\ref{fig:hll}. This algorithm is able to estimate cardinalities greater than $10^9$ with a typical standard error of 4\%, using only 1.5 kB of memory. Crucially, the proof of this~\cite[Theorem 1]{hll} makes the assumption that the inputs are drawn independently and uniformly at random from the input domain -- loosely, it is an average case analysis.\footnote{Technically, the results of~\cite{hll} are proven for a distribution obtained by making arbitrary replications and permutations after sampling independently and uniformly, to simulate input taken from an \emph{ideal multiset}. But they also hold without replication and permutation because HLL's internal state is invariant under repeated inputs.}

@@

We first begin by explaining the high-level idea of our approach. We allow an adversary ${\mathcal{A}}$ to interact with an HLL via various oracles that capture its capabilities in each of the three scenarios S2, S3 and S4: it may have an \texttt{insert($\cdot$)} oracle allowing it to add items to the HLL sketch, a \texttt{read} oracle that returns the HLL cardinality estimate, and a \texttt{corrupt} that returns all the HLL bucket contents (but not any secret key $K$); in S4 it has all three oracles. In addition, in all 3 scenarios, ${\mathcal{A}}$ has access to an \texttt{init($\cdot$)} oracle by which it sets the HLL parameters $n$, $m=2^n$, $\ell$, $\alpha_m$ used to initialise the HLL sketch. We assume ${\mathcal{A}}$ always makes exactly one \texttt{init($\cdot$)} query and that it is its first oracle call.

In the security model, the game selects at random one of two worlds by sampling a bit $b$. In the real world ($b=0$), ${\mathcal{A}}$ interacts with the scheme HLL-$F$. This is the same as HLL operating as in Figure~\ref{fig:hll} but with the hash function $h$ replaced by a keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$ whose key $K$ is sampled uniformly at random at the beginning of the game and not available to the adversary; ${\mathcal{A}}$ has access to the relevant selection of oracles and completes its execution with some output $\textsf{out}_{\mathcal{A}}$ (a bit-string).

@@

We will prove that  ${\mathcal{S}}$ is such that, given any adversary ${\mathcal{A}}$, then any distinguisher $\mathcal{E}$ that can distinguish between the $\textsf{out}_{\mathcal{A}}$ in the real world and $\textsf{out}_{\mathcal{S}}$ in the ideal world can be used to build an adversary ${{\mathcal{B}}}$ which breaks the PRF security of $F$. Under the assumption that $F$ is a good PRF, no such efficient distinguisher $\mathcal{E}$ can exist, and we may conclude that (up to PRF security) no efficient distinguisher can tell that it is interacting with the ideal world instead of the real world.

Why does this proof imply that ${\mathcal{A}}$ cannot manipulate the cardinality estimate? To see this, note that (amongst many other possibilities) ${\mathcal{A}}$'s output could be the cardinality estimate obtained from a \texttt{read} query, allowing a specific ${\mathcal{E}}$ to try to use that to distinguish between the two worlds. Yet, by the preceding argument, the simulator in the ideal world is able to produce an output that ${\mathcal{E}}$ cannot distinguish from that of ${\mathcal{A}}$ in the real world. In particular, our simulator will output whatever ${\mathcal{A}}$ outputs, so if ${\mathcal{A}}$'s output $\textsf{out}_{\mathcal{A}}$ is the result of a \texttt{read} query in the ideal world, then so will ${\mathcal{S}}$'s. But in the ideal world, ${\mathcal{S}}$ uses random sampling for the $y$ values, so in that world ${\mathcal{S}}$ (and ${\mathcal{A}}$ that it runs) are operating in the setting where the average case analysis of HLL applies, as per~\cite{hll}. So in the ideal world this specific ${\mathcal{E}}$ would receive an accurate average case cardinality estimate. Since ${\mathcal{E}}$ tries to distinguish the two worlds from the cardinality estimate and we already showed that no efficient distinguisher ${{\mathcal{E}}}$ can exist, it must follow that the cardinality estimates produced in the real and ideal worlds are indistinguishable (up to PRF security). It then follows that ${\mathcal{A}}$ has no advantage when it comes to manipulating the HLL-$F$ cardinality estimate. So HLL, when instantiated with PRF $F$, is secure against cardinality manipulation attacks.
% in S4 (and therefore in the weaker attacks models S2 and S3 as well).

@@

We now formalise the above intuition. We consider distinguisher ${\mathcal{E}}$ to be an algorithm running in a security game. The game begins with the selection of a bit $b$ uniformly at random and then runs either ${\mathcal{A}}$ (the real world, $b=0$) or ${\mathcal{S}}$ (the ideal world, $b=1$). Adversary ${\mathcal{A}}$ interacts with up to four oracles which are handled by the game using the algorithms of HLL-$F$; different combinations of oracles define scenarios S2, S3, S4. After making some queries, ${\mathcal{A}}$ returns its output $\textsf{out}_{\mathcal{A}}$ to ${\mathcal{E}}$. Simulator ${\mathcal{S}}$ runs ${\mathcal{A}}$ in a blackbox manner (as specified below) and receives ${\mathcal{A}}$'s output. It then provides its own output $\textsf{out}_{\mathcal{S}}$ to ${\mathcal{E}}$. Finally, ${\mathcal{E}}$ outputs a bit $b'$ estimating whether it is interacting with the real world ($b'=0$) or the ideal world ($b'=1$). For $i=2, 3, 4$, we define ${\mathcal{E}}$'s S$i$-Real-or-Ideal advantage against HLL-$F$ with adversary ${\mathcal{A}}$ and simulator ${\mathcal{S}}$ to be:
\[
\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny{HLL-$F$, ${\mathcal{A}}$, ${\mathcal{S}}$}}}({\mathcal{E}}) := \lvert 2\Pr[b' = b] - 1 \rvert \, .
\]
%Using a standard transformation, this probability can be rewritten as $\lvert \Pr[b' = 1| b = 1] - \Pr[b'=0|b=1] \rvert$.

The four oracles that  ${\mathcal{A}}$ can interact with behave as follows in the real world:
\begin{itemize}
\item \texttt{init($\cdot$)}: this oracle receives as input the HLL-$F$ parameters ($n$, $\ell$, $\alpha_m$) and performs initialisation of the HLL-$F$ sketch, including generating the array of $m=2^n$ buckets $M[j]$ and initialising all the entries to $0$, and selecting a key $K \leftarrow_{\$} \{0,1\}^k$ for keyed function $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$. This oracle has no output. It is available in all of S2, S3, S4. We assume it is only called once as ${\mathcal{A}}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$, the oracle implements HLL-$F$ by setting $y = F_K(x)$ and following the procedure in Figure~\ref{fig:hll} to insert $x$ into the HLL sketch (that is, it uses the first $n$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$). This oracle has no output. It is available in all of S2, S3, S4. Since HLL and variants of it we consider here have the property that reinserting an item $x$ does not change the internal state of the HLL sketch (nor the cardinality estimate computed from it) we assume without loss of generality that all queries $x$ made by ${\mathcal{A}}$ to \texttt{insert($\cdot$)} are distinct.
\item \texttt{read}: this oracle has no input and returns the HLL-$F$ cardinality estimate as per Figure~\ref{fig:hll}, i.e.\ it computes a raw cardinality estimate via $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$ and then performs corrections for small and large values. We allow multiple queries to this oracle and make it available to ${\mathcal{A}}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle has no input and returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; here we allow multiple queries because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

\begin{definition}
We say that the scheme HLL-$F$ is S$i$-simulation-secure with parameters $(q,t_{\mathcal{S}},t_{\mathcal{A}},t_{\mathcal{E}},\epsilon)$ if there exists a simulator $\mathcal{S}$ running in time at most $t_{\mathcal{S}}$ such that, for all adversaries ${\mathcal{A}}$ running in time at most $t_{\mathcal{A}}$ and making at most $q$ \texttt{insert($\cdot$)} queries and all distinguishers ${\mathcal{E}}$ running in time at most $t_{\mathcal{E}}$, we have
$\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny{HLL-$F$, ${\mathcal{A}}$, ${\mathcal{S}}$}}}({\mathcal{E}}) \leq \epsilon$. 
\end{definition}

@@

Note that our security analysis using a simulation-based definition does not require us to specify a winning condition for adversary $\mathcal{A}$. Rather, we infer from inspecting the simulator what the adversary can (or cannot) achieve in the ideal world, and show that no efficient algorithm $\mathcal{E}$ can distinguish the real and ideal worlds. This is in contrast to the game-based approach of~\cite{CCS:ClaPatShrPS19} which needs a complex definition to say when the adversary has successfully manipulated a probabilistic data structure. This reveals a strength of the simulation-based approach in more complex settings like ours.

\vspace{0.1cm}
\noindent\textbf{A specific simulator ${\mathcal{S}}$.}
We next describe the simulator ${\mathcal{S}}$ that we will use in our further analysis: ${\mathcal{S}}$ operates by running ${\mathcal{A}}$ in a blackbox manner and handling all of its queries as follows: 
\begin{itemize}
\item \texttt{init($\cdot$)}: on input $n$, $\ell$, $\alpha_m$, ${\mathcal{S}}$ initialises an HLL sketch as in Figure~\ref{fig:hll}; that is, it creates an array of $m=2^n$ buckets $M[j]$ and initialises them all to 0.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{D}}}$,  ${\mathcal{S}}$ ignores $x$, sets $y \leftarrow_{\$} \{0,1\}^{n+\ell}$, and proceeds to update the buckets using this value of $y$ as in the original HLL algorithm in Figure~\ref{fig:hll}.
\item \texttt{read}:  ${\mathcal{S}}$ returns the HLL cardinality estimate computed as per Figure~\ref{fig:hll}.
\item \texttt{corrupt}: ${\mathcal{S}}$ returns the current vector $(M[j])_{j=1}^{m}$ of HLL bucket contents that it has stored.
\end{itemize}
At the end of ${\mathcal{A}}$'s execution, ${\mathcal{S}}$ takes ${\mathcal{A}}$'s output and forwards it to ${\mathcal{E}}$ as its own output. Notice that the running time of  ${\mathcal{S}}$ is (roughly) equal to that of ${\mathcal{A}}$.

For our definition to meaningfully capture security against cardinality manipulation, it is important to restrict ${\mathcal{S}}$ to doing random sampling when handling \texttt{insert($\cdot$)} queries. Otherwise, our definition could be vacuously satisfied (for example ${\mathcal{S}}$ could just sample the PRF key $K$ for itself and provide a perfect real-world view). This is somewhat different to common simulation-based security definitions, where ${\mathcal{S}}$'s input might be restricted (e.g.\ in a confidentiality definition, ${\mathcal{S}}$ might receive only $|m|$ in place of $m$) but its behaviour is not. Crucially, in our setting, security arises from inspecting the details of ${\mathcal{S}}$ and observing that it instantiates HLL in the average case.

We require one further standard definition.
\begin{definition}
Let $F: \{0,1\}^k \times {{\mathcal{D}}} \rightarrow  \{0,1\}^{n+\ell}$.
We define the PRF advantage of an adversary ${{\mathcal{B}}}$ against $F$ outputting a bit $d'$ as being its advantage in distinguishing whether it is interacting with $F$ for a uniformly random key $K$ ($d=0$) or interacting with a truly random function ($d=1$). We write: 
\[
\mbox{Adv}^{\mbox{\tiny PRF}}_{F}({{\mathcal{B}}}) := \lvert 2\Pr[d' = d] - 1 \rvert\,.
\]
\end{definition}
Here, we assume without loss of generality that all of ${{\mathcal{B}}}$'s queries are distinct. See~\cite[Section 4.4.1]{BonehShoup} for details of the PRF security definition.

Now we are ready to formally state our main result. The proof can be found in Appendix~\ref{sec:proof}.
\begin{theorem}\label{thm:security-of-HLL-with-PRF}
Let simulator $\mathcal{S}$ be as described above. For any S4 adversary ${\mathcal{A}}$ against HLL-$F$ and any distinguisher $\mathcal{E}$, we can construct a PRF adversary ${{\mathcal{B}}}$ against $F$ such that:
\[
 \mbox{Adv}^{\mbox{\tiny PRF}}_{F}({{\mathcal{B}}}) = 
 \mbox{Adv}^{\mbox{\tiny S$4$-RoI}}_{\mbox{\tiny{HLL-$F$, ${\mathcal{A}}$, ${\mathcal{S}}$}}}({\mathcal{E}}) \,.
\]
Moreover, if ${\mathcal{A}}$ makes $q$ queries to its \texttt{insert($\cdot$)} oracle, then ${{\mathcal{B}}}$ makes $q$ queries to its real or random oracle. Finally, the running time of ${{\mathcal{B}}}$ is (roughly) $t_{\mathcal{A}}+t_{\mathcal{E}}$.
\end{theorem}

>>>>

\begin{figure*}[tb!]
\centering
     \makebox[\linewidth]{
\fbox{
\begin{minipage}{40em}
\begin{flushleft}
\textit{Let $h : \mathcal{Y} \rightarrow \{0, 1\}^{32}$ hash data from some domain $\mathcal{Y}$ to binary $32$-bit strings.\\
For $s$ of bit-length $\ell$, let $\rho(s)$ denote the position of the leftmost $1$-bit of $s$: e.g., $\rho(1...) = 1, \rho(0001...) = 4, \rho(0^\ell) = \ell + 1.$\\}
\textbf{define} $m = 2^n$ with $n \in [4..16].$\\
\textbf{define} $\alpha_{16} = 0.673$; $\alpha_{32} = 0.697$; $\alpha_{64} = 0.709$; $\alpha_{m} = 0.7213/(1+ 1.079/m)$ for $m \ge 128$;\\
\vspace{0.5em}
\textbf{Algorithm} HyperLogLog (\textbf{input} ${\mathcal{X}}$ : multiset of items from domain $\mathcal{Y}$).\\
\textbf{initialize} a collection of $m$ registers, $M[1], ..., M[m]$, to $0$;\\
\vspace{0.5em}
\textbf{for} $x \in {\mathcal{X}}$ \textbf{do}\\
    \hspace{2em} \textbf{set} $y := h(x);$\\
    \hspace{2em} \textbf{set} $j = 1 + <y_1y_2...y_{n}>_2$; \hfill \textit{\{bucket index determined by the first $n$ bits of $y$\}}\\
    \hspace{2em} \textbf{set} $w := y_{n+1}y_{n+2} \ldots y_{n+\ell}$;\hfill \textit{\{bit string determined by the last $\ell$ bits of $y$\}}\\
   \hspace{2em} \textbf{set} $M[j] :=$ max$(M[j], \rho(w))$;\\
\vspace{0.2em}
\textbf{compute} $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$; \hfill \textit{\{the raw HyperLogLog estimate\}}\\
\vspace{0.2em}
\textbf{if} $E \le \frac{5}{2}m$ \textbf{then}\\
    \hspace{2em}  \textbf{let} $V$ be the number of registers equal to $0$;\\
    \hspace{2em}  \textbf{if} $V \neq 0$ \textbf{then set} $E^*:=  m\cdot \ln(m/V)$ \textbf{else set} $E^*:= E$; \hfill \textit{\{small range correction\}}\\
\textbf{if} $E \le \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em} \textbf{set} $E^*:= E$; \hfill \textit{\{intermediate range --- no correction\}}\\
\textbf{if} $E > \frac{1}{30}2^{32}$ \textbf{then}\\
    \hspace{2em}  \textbf{set} $E^*:= -2^{32}$ln$(1 - E/2^{32})$; \hfill \textit{\{large range correction\}}\\
\textbf{return} \textit{cardinality estimate $E^*$ with typical relative error} $\pm1.04/\sqrt{m}$.
\end{flushleft}
\end{minipage}
}}
\caption{The HyperLogLog Algorithm from~\cite{hll}}
\label{fig:hll}
\end{figure*}

@@

In the original HLL proposal~\cite{hll}, relatively small and large cardinalities are handled separately to correct for systematic errors. When the raw cardinality estimate is below $\frac{5}{2}m$, the Hit Counting algorithm of Whang \emph{et al.}~\cite{hitcounting} is used to produce the final cardinality estimate. This estimator uses the number of empty buckets (whose register value is 0) instead of the number of leading zeros in the stream. A similar correction is applied for large cardinalities in order to take into account the likelihood of hash collisions. In addition, the cardinality estimate is usually rounded to a whole number before being output. The original HLL algorithm from~\cite{hll} is shown in Figure~\ref{fig:hll}. This algorithm is able to estimate cardinalities greater than $10^9$ with a typical standard error of 4\%, using only 1.5 kB of memory. Crucially, the proof of this~\cite[Theorem 1]{hll} makes the assumption that the inputs are drawn independently and uniformly at random from the input domain -- loosely put, it is an average case analysis.\footnote{Technically, the results of~\cite{hll} are proven for a distribution obtained by making arbitrary replications and permutations after sampling independently and uniformly, to simulate input taken from an \emph{ideal multiset}. But they also hold without replication and permutation because HLL's internal state is invariant under repeated inputs.}

@@

We first begin by explaining the high-level idea of our approach. We allow an adversary ${\mathcal{A}}$ to interact with an HLL sketch via various oracles that capture its capabilities in each of the three scenarios S2, S3 and S4: it may have an \texttt{insert($\cdot$)} oracle allowing it to add items to the HLL sketch, a \texttt{read} oracle that returns the HLL cardinality estimate, and a \texttt{corrupt} that returns all the HLL sketch's bucket contents (but not any secret key $K$); in S4 it has all three oracles. In addition, in all 3 scenarios, ${\mathcal{A}}$ has access to an \texttt{init($\cdot$)} oracle by which it sets the HLL parameters $n$, $m=2^n$, $\ell$, $\alpha_m$ used to initialise the HLL sketch. We assume ${\mathcal{A}}$ always makes exactly one \texttt{init($\cdot$)} query and that it is its first oracle call.

In the security model, the game selects at random one of two worlds by sampling a bit $b$. In the real world ($b=0$), ${\mathcal{A}}$ interacts with the scheme HLL-$F$. This is the same as HLL operating as in Figure~\ref{fig:hll} but where the hash function $h$ is replaced by a keyed function $F: \{0,1\}^k \times {{\mathcal{Y}}} \rightarrow  \{0,1\}^{n+\ell}$ whose key $K$ is sampled uniformly at random at the beginning of the game and not available to the adversary; ${\mathcal{A}}$ has access to the relevant selection of oracles and completes its execution with some output $\textsf{out}_{\mathcal{A}}$ (a bit-string).

@@

We will prove that  ${\mathcal{S}}$ is such that, given any adversary ${\mathcal{A}}$, then any distinguisher $\mathcal{D}$ that can distinguish between the $\textsf{out}_{\mathcal{A}}$ in the real world and $\textsf{out}_{\mathcal{S}}$ in the ideal world can be used to build an adversary ${{\mathcal{B}}}$ which breaks the PRF security of $F$. Under the assumption that $F$ is a good PRF, no such efficient distinguisher $\mathcal{D}$ can exist, and we may conclude that (up to PRF security) no efficient distinguisher can tell that it is interacting with the ideal world instead of the real world.

Why does this proof imply that ${\mathcal{A}}$ cannot manipulate the cardinality estimate? To see this, note that (amongst many other possibilities) ${\mathcal{A}}$'s output could be the cardinality estimate obtained from a \texttt{read} query, allowing a specific ${\mathcal{D}}$ to try to use that to distinguish between the two worlds. Yet, by the preceding argument, the simulator in the ideal world is able to produce an output that ${\mathcal{D}}$ cannot distinguish from that of ${\mathcal{A}}$ in the real world. In particular, our simulator will output whatever ${\mathcal{A}}$ outputs, so if ${\mathcal{A}}$'s output $\textsf{out}_{\mathcal{A}}$ is the result of a \texttt{read} query in the ideal world, then so will ${\mathcal{S}}$'s. But in the ideal world, ${\mathcal{S}}$ uses random sampling for the $y$ values, so in that world ${\mathcal{S}}$ (and ${\mathcal{A}}$ that it runs) are operating in the setting where the average case analysis of HLL applies, as per~\cite{hll}. So in the ideal world this specific ${\mathcal{D}}$ would receive an accurate average case cardinality estimate. Since ${\mathcal{D}}$ tries to distinguish the two worlds from the cardinality estimate and we already showed that no efficient distinguisher ${{\mathcal{D}}}$ can exist, it must follow that the cardinality estimates produced in the real and ideal worlds are indistinguishable (up to PRF security). It then follows that ${\mathcal{A}}$ has no advantage when it comes to manipulating the HLL-$F$ cardinality estimate. So HLL, when instantiated with PRF $F$, is secure against cardinality manipulation attacks.
% in S4 (and therefore in the weaker attacks models S2 and S3 as well).

@@

We now formalise the above intuition. Our definitions are specific to HLL-$F$ as the object of analysis, but are easily generalised to any cardinality estimator with the same syntax as HLL-$F$ (formally defined by a triple of algorithms (\texttt{Init}, \texttt{Insert}, \texttt{Read})).

\vspace{0.2cm}
\noindent\textbf{The distinguisher ${\mathcal{D}}$ and its advantage.} We consider distinguisher ${\mathcal{D}}$ to be an algorithm running in a security game. The game begins with the selection of a bit $b$ uniformly at random and then runs either adversary ${\mathcal{A}}$ (the real world, $b=0$) or simulator ${\mathcal{S}}$ (the ideal world, $b=1$). Adversary ${\mathcal{A}}$ interacts with up to four oracles which are handled by the game using the algorithms of HLL-$F$. These are defined below; different combinations of oracles define scenarios S2, S3, S4. After making some queries, ${\mathcal{A}}$ returns its output $\textsf{out}_{\mathcal{A}}$ to ${\mathcal{D}}$. Simulator ${\mathcal{S}}$ runs ${\mathcal{A}}$ in a blackbox manner (as specified below) and receives ${\mathcal{A}}$'s output. It then provides its own output $\textsf{out}_{\mathcal{S}}$ to ${\mathcal{D}}$. Finally, ${\mathcal{D}}$ outputs a bit $b'$ estimating whether it is interacting with the real world ($b'=0$) or the ideal world ($b'=1$). For $i=2, 3, 4$, we define ${\mathcal{D}}$'s S$i$-Real-or-Ideal advantage against HLL-$F$ with adversary ${\mathcal{A}}$ and simulator ${\mathcal{S}}$ to be:
\[
\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny{HLL-$F$,${\mathcal{A}}$,${\mathcal{S}}$}}}({\mathcal{D}}) := \lvert 2\Pr[b' = b] - 1 \rvert \, .
\]
%Using a standard transformation, this probability can be rewritten as $\lvert \Pr[b' = 1| b = 1] - \Pr[b'=0|b=1] \rvert$.

\vspace{0.2cm}
\noindent\textbf{The adversary ${\mathcal{A}}$ and its oracles.}
The four oracles that  ${\mathcal{A}}$ can interact with behave as follows in the real world:
\begin{itemize}
\item \texttt{init($\cdot$)}: this oracle receives as input the HLL-$F$ parameters ($n$, $\ell$, $\alpha_m$) and performs initialisation of the HLL-$F$ sketch, including generating the array of $m=2^n$ buckets $M[j]$ and initialising all the entries to $0$, and selecting a key $K \leftarrow_{\$} \{0,1\}^k$ for keyed function $F: \{0,1\}^k \times {{\mathcal{Y}}} \rightarrow  \{0,1\}^{n+\ell}$. This oracle has no output. It is available in all of S2, S3, S4. We assume it is only called once as ${\mathcal{A}}$'s first oracle query.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{Y}}}$, the oracle implements HLL-$F$ by setting $y = F_K(x)$ and following the procedure in Figure~\ref{fig:hll} to insert $x$ into the HLL sketch (that is, it uses the first $n$ bits of $y$ to determine a bucket index $j$ and the next $\ell$ bits of $y$ to determine a bit-string $w$; then $\rho(w)$ is used to update the bucket at index $j$ via $M[j]:= \max(M[j],\rho(w))$). This oracle has no output. It is available in all of S2, S3, S4. Since HLL and variants of it we consider here have the property that reinserting an item $x$ does not change the internal state of the HLL sketch (nor the cardinality estimate computed from it) we assume without loss of generality that all queries $x$ made by ${\mathcal{A}}$ to \texttt{insert($\cdot$)} are distinct.
\item \texttt{read}: this oracle has no input and returns the HLL-$F$ cardinality estimate as per Figure~\ref{fig:hll}, i.e.\ it computes a raw cardinality estimate via $E := \alpha_mm^2\cdot(\sum_{j=1}^m2^{-M[j]})^{-1}$ and then performs corrections for small and large values. We allow multiple queries to this oracle and make it available to ${\mathcal{A}}$ in S3 and S4, but not S2.
\item \texttt{corrupt}: this oracle has no input and returns the vector $(M[j])_{j=1}^{m}$ of HLL bucket contents (but not any secret key $K$). This oracle is available only in S4. We allow the adversary to query \texttt{corrupt} as often as it wishes during its attack. Our description of S4 in Section~\ref{sec:attacks} limited the adversary to at most one such query; here we allow multiple queries because doing so does not increase the difficulty of achieving security.  
\end{itemize}

@@

\vspace{0.2cm}
\noindent\textbf{A specific simulator ${\mathcal{S}}$.} In order to obtain a meaningful security definition that prevents cardinality manipulation, we need to restrict ${\mathcal{S}}$ to doing random sampling when handling \texttt{insert($\cdot$)} queries coming from ${\mathcal{A}}$. Otherwise, our definition could be vacuously satisfied (for example ${\mathcal{S}}$ could just sample the PRF key $K$ for itself and provide a perfect real-world view). This is somewhat different to typical simulation-based security definitions, where ${\mathcal{S}}$'s behaviour is not restricted but its input might be (e.g.\ in a confidentiality definition, ${\mathcal{S}}$ might receive only $|m|$ in place of plaintext $m$). Crucially, in our setting, security arises from inspecting the details of ${\mathcal{S}}$ and observing that, due to its use of random sampling, it instantiates HLL in the average case.

In full detail, our specific ${\mathcal{S}}$ operates by running ${\mathcal{A}}$ in a blackbox manner and handling all of its queries as follows:
\begin{itemize}
\item \texttt{init($\cdot$)}: on input $n$, $\ell$, $\alpha_m$, ${\mathcal{S}}$ initialises an HLL sketch as in Figure~\ref{fig:hll}; that is, it creates an array of $m=2^n$ buckets $M[j]$ and initialises them all to 0.
\item \texttt{insert($\cdot$)}: on input $x \in {{\mathcal{Y}}}$,  ${\mathcal{S}}$ ignores $x$, sets $y \leftarrow_{\$} \{0,1\}^{n+\ell}$, and proceeds to update the buckets using this value of $y$ as in the original HLL algorithm in Figure~\ref{fig:hll}.
\item \texttt{read}:  ${\mathcal{S}}$ returns the HLL cardinality estimate computed as per Figure~\ref{fig:hll}.
\item \texttt{corrupt}: ${\mathcal{S}}$ returns the current vector $(M[j])_{j=1}^{m}$ of HLL bucket contents that it has stored.
\end{itemize}
At the end of ${\mathcal{A}}$'s execution, ${\mathcal{S}}$ takes ${\mathcal{A}}$'s output $\textsf{out}_{\mathcal{A}}$ and forwards it to ${\mathcal{D}}$ as its own output $\textsf{out}_{\mathcal{S}}$. Notice that the running time $t_{\mathcal{S}}$ of ${\mathcal{S}}$ is (roughly) equal to that of ${\mathcal{A}}$.

\vspace{0.2cm}
\noindent\textbf{Formal security definition and main result.} Now we are ready to give:
\begin{definition}
The scheme HLL-$F$ is S$i$-simulation-secure with parameters $(q,t_{\mathcal{S}},t_{\mathcal{A}},t_{\mathcal{D}},\epsilon)$ if, for the specific simulator $\mathcal{S}$ given above, for all adversaries ${\mathcal{A}}$ running in time at most $t_{\mathcal{A}}$ and making at most $q$ \texttt{insert($\cdot$)} queries, and for all distinguishers ${\mathcal{D}}$ running in time at most $t_{\mathcal{D}}$, we have
$\mbox{Adv}^{\mbox{\tiny S$i$-RoI}}_{\mbox{\tiny{HLL-$F$,${\mathcal{A}}$,${\mathcal{S}}$}}}({\mathcal{D}}) \leq \epsilon$. 
\end{definition}

@@

Note that our security analysis using a simulation-based definition does not require us to specify a winning condition for adversary $\mathcal{A}$. Rather, we infer from inspecting the simulator what the adversary can (or cannot) achieve in the ideal world, and show that no efficient algorithm $\mathcal{D}$ can distinguish the real and ideal worlds. This is in contrast to the game-based approach of~\cite{CCS:ClaPatShrPS19} which needs a complex definition to say when the adversary has successfully manipulated a probabilistic data structure. This reveals a strength of the simulation-based approach.

Note also that our definition is given with respect to a \emph{specific} simulator, but could be generalised to allow any simulator that performs random sampling when handling \texttt{insert($\cdot$)} queries.

Before stating our main result, we require one further standard definition:
\begin{definition}
Let $F: \{0,1\}^k \times {{\mathcal{Y}}} \rightarrow  \{0,1\}^{n+\ell}$.
We define the PRF advantage of an adversary ${{\mathcal{B}}}$ against $F$ outputting a bit $d'$ as being its advantage in distinguishing whether it is interacting with $F$ for a uniformly random key $K$ ($d=0$) or interacting with a truly random function ($d=1$). We write: 
\[
\mbox{Adv}^{\mbox{\tiny PRF}}_{F}({{\mathcal{B}}}) := \lvert 2\Pr[d' = d] - 1 \rvert\,.
\]
\end{definition}
Here, we assume without loss of generality that all of ${{\mathcal{B}}}$'s queries are distinct. See~\cite[Section 4.4.1]{BonehShoup} for details of the PRF security definition.

Now we are ready to formally state our main result. The proof can be found in Appendix~\ref{sec:proof}.
\begin{theorem}\label{thm:security-of-HLL-with-PRF}
Let simulator $\mathcal{S}$ be as described above. For any S4 adversary ${\mathcal{A}}$ against HLL-$F$ and any distinguisher $\mathcal{D}$, we can construct a PRF adversary ${{\mathcal{B}}}$ against $F$ such that:
\[
 \mbox{Adv}^{\mbox{\tiny PRF}}_{F}({{\mathcal{B}}}) = 
 \mbox{Adv}^{\mbox{\tiny S$4$-RoI}}_{\mbox{\tiny{HLL-$F$,${\mathcal{A}}$,${\mathcal{S}}$}}}({\mathcal{D}}) \,.
\]
Moreover, if ${\mathcal{A}}$ makes $q$ queries to its \texttt{insert($\cdot$)} oracle, then ${{\mathcal{B}}}$ makes $q$ queries to its real or random oracle. Finally, the running time of ${{\mathcal{B}}}$ is (roughly) $t_{\mathcal{A}}+t_{\mathcal{D}}$.
\end{theorem}
